---
title: "Effect of Enhanced Seasonal Malaria Chemoprevention on Hemoglobin Trajectories in Children"
subtitle: "Introduction to Data Analysis - Group 1 Project Code"
published-title: "Submitted On"
author:
  - "Your Name"
date: today
abstract: ""

format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    theme: cosmo
    embed-resources: true
    html-math-method: mathjax
    self-contained: true
    self-contained-math: false
    include-in-header:
      text: |
        <style>
        body {
          text-align: justify;
          text-justify: inter-word;
          -webkit-user-select: text;
          -moz-user-select: text;
          -ms-user-select: text;
          user-select: text;
        }

        table, th, td, caption {
          -webkit-user-select: text !important;
          -moz-user-select: text !important;
          -ms-user-select: text !important;
          user-select: text !important;
        }

        img, figure {
          pointer-events: auto !important;
          -webkit-user-drag: auto;
          user-select: auto !important;
        }
        </style>

        <script>
        document.addEventListener("contextmenu", function(e) {
          e.stopPropagation();
        }, true);
        </script>
  pdf:
    toc: true
    number-sections: true

execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(performance)
library(finalfit)
library(knitr)
library(kableExtra)
library(ggpubr)
library(tibble)

pal_trt <- c("Control" = "brown", "eSMC" = "black")
```


```{r}
#| label: data-import

df_raw <- read.csv("malaria_anemia_rct.csv")

df <- df_raw %>%
  mutate(
    child_id = factor(child_id),
    treatment = factor(treatment, levels = c(0, 1),
                       labels = c("Control", "eSMC")),
    itn_use = factor(itn_use, levels = c(0, 1),
                     labels = c("No", "Yes")),
    month = as.numeric(month)
  ) %>%
  arrange(child_id, month)

glimpse(df)
summary(df)
```

```{r cleaning}
#| include: true

hb_range  <- c(4, 20)
age_range <- c(6, 120)
valid_months <- c(0, 2, 4, 6)

flag_hb  <- df |> filter(!between(hemoglobin, hb_range[1], hb_range[2]))
flag_age <- df |> filter(!between(age_baseline, age_range[1], age_range[2]))
flag_mon <- df |> filter(!month %in% valid_months)

cat("Flagged Hb outliers :", nrow(flag_hb), "\n")
cat("Flagged age outliers:", nrow(flag_age), "\n")
cat("Invalid month values:", nrow(flag_mon), "\n")

dupl <- df |> group_by(child_id, month) |> filter(n() > 1)
cat("Duplicate child×month rows:", nrow(dupl), "\n")
```

---

# Results

## Baseline Characteristics of the Cohort

@tbl-Baseline summarises the baseline characteristics of the study cohort stratified by treatment group.

```{r}
#| label: tbl-Baseline
#| tbl-cap: "Baseline characteristics by treatment group"

df_baseline <- df |>
  filter(month == 0)

explanatory <- c("age_baseline", "itn_use", "hemoglobin")
dependent <- "treatment"

baseline_table <- df_baseline %>%
  summary_factorlist(
    dependent,
    explanatory,
    p = TRUE,
    p_cont_para = "t.test",
    p_cat = "chisq",
    add_dependent_label = TRUE,
    cont = "mean",
    total_col = FALSE,
    add_col_totals = TRUE,
    add_row_totals = TRUE,
    na_include = FALSE,
    na_to_p = FALSE,
    na_to_prop = FALSE,
    include_col_totals_percent = FALSE
  )

names(baseline_table)[1] <- "Variable"

baseline_table[, 1] <- dplyr::recode(
  baseline_table[, 1],
  "age_baseline" = "Baseline Age",
  "itn_use" = "ITN Use",
  "hemoglobin" = "Hemoglobin"
)

baseline_table %>%
  kable(
    caption = "",
    format = "html",
    booktabs = TRUE,
    escape = FALSE,
    align = c("l", "c", "c", "l", "c", "c", "c")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "bordered"),
    full_width = FALSE,
    font_size = 16,
    position = "center"
  ) %>%
  row_spec(
    0,
    bold = TRUE,
    background = "#003366",
    color = "white",
    font_size = 18
  ) %>%
  row_spec(
    1:nrow(baseline_table),
    color = "black",
    background = "white"
  ) %>%
  column_spec(
    1,
    bold = TRUE,
    color = "black",
    width = "12em",
    extra_css = "white-space: nowrap;"
  ) %>%
  column_spec(2, color = "black", width = "9em") %>%
  column_spec(3, color = "black", width = "8em") %>%
  column_spec(4, color = "black", width = "10em") %>%
  column_spec(5, color = "black", width = "8em") %>%
  column_spec(6, color = "black", width = "8em") %>%
  column_spec(7, color = "black", width = "5em")
```

```{r}
#| label: fig-base
#| fig-cap: "Baseline characteristics by treatment group shown as grouped bars."

cont_df <- df_baseline |>
  group_by(treatment) |>
  summarise(
    `Baseline Age (months)` = mean(age_baseline, na.rm = TRUE),
    `Haemoglobin (g/dL)` = mean(hemoglobin, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = -treatment,
    names_to = "Characteristic",
    values_to = "Value"
  ) |>
  mutate(Panel = "Mean values")

cat_df <- df_baseline |>
  count(treatment, itn_use) |>
  group_by(treatment) |>
  mutate(Value = 100 * n / sum(n)) |>
  ungroup() |>
  mutate(
    Characteristic = paste("ITN Use:", itn_use),
    Panel = "Percent"
  ) |>
  select(treatment, Characteristic, Value, Panel)

plot_df <- bind_rows(cont_df, cat_df)

ggplot(plot_df, aes(x = Characteristic, y = Value, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_text(
    aes(label = round(Value, 1)),
    position = position_dodge(width = 0.75),
    vjust = -0.3,
    size = 3
  ) +
  facet_grid(. ~ Panel, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = pal_trt, name = "Treatment") +
  labs(x = "Baseline characteristics", y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
#| label: fig-box
#| fig-cap: "Boxplot of haemoglobin by visit and treatment groups."

df |>
  mutate(month_f = factor(month)) |>
  ggplot(aes(x = month_f, y = hemoglobin, fill = treatment)) +
  annotate(
    "rect",
    xmin = -Inf, xmax = Inf,
    ymin = 4, ymax = 20,
    fill = "gray", alpha = 0.25
  ) +
  geom_boxplot(
    width = 0.6,
    alpha = 0.7,
    outlier.colour = "red",
    outlier.size = 1.8
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 3,
    color = "black"
  ) +
  scale_fill_manual(values = pal_trt, name = "Treatment") +
  geom_hline(
    yintercept = c(4, 20),
    linetype = "dashed",
    colour = "darkgreen",
    linewidth = 0.5
  ) +
  geom_hline(
    yintercept = 11,
    linetype = "dotted",
    colour = "#8B0000",
    linewidth = 0.7
  ) +
  labs(
    title = "Haemoglobin by Visit and Treatment Group",
    x = "Month",
    y = "Haemoglobin (g/dL)"
  ) +
  theme_minimal()
```


## Haemoglobin Trajectories Over Time

The individual and group-level haemoglobin patterns are shown in Figure @fig-traj and Figure @fig-meantraj.



```{r}
library(dplyr)
library(ggplot2)
library(ggpubr)
library(ggnewscale)

set.seed(42)

sampled_ids <- sample(levels(df$child_id), 300)
df_samp <- df |> filter(child_id %in% sampled_ids)

pal_trt_traj <- c("Control" = "black", "eSMC" = "darkgreen")

pal_overlay <- c("Control" = "lightblue", "eSMC" = "black")

df_means <- df |>
  group_by(treatment, month) |>
  summarise(
    mean_hb = mean(hemoglobin, na.rm = TRUE),
    se_hb   = sd(hemoglobin, na.rm = TRUE) / sqrt(sum(!is.na(hemoglobin))),
    .groups = "drop"
  )

ggplot(df_samp, aes(x = month, y = hemoglobin,
                    group = child_id, colour = treatment)) +
  geom_line(alpha = 0.8, linewidth = 0.4) +
  geom_point(alpha = 0.25, size = 0.7) +
  scale_colour_manual(values = pal_trt_traj, guide = "none") +
  ggnewscale::new_scale_color() +
  geom_ribbon(
    data = df_means,
    aes(
      x = month,
      ymin = mean_hb - 1.96 * se_hb,
      ymax = mean_hb + 1.96 * se_hb,
      fill = treatment,
      group = treatment
    ),
    inherit.aes = FALSE,
    alpha = 0.25,
    colour = NA
  ) +
  geom_line(
    data = df_means,
    aes(x = month, y = mean_hb, group = treatment, colour = treatment),
    inherit.aes = FALSE,
    linewidth = 2
  ) +
  geom_point(
    data = df_means,
    aes(x = month, y = mean_hb, colour = treatment),
    inherit.aes = FALSE,
    size = 4
  ) +
  scale_color_manual(values = pal_overlay, name = "Treatment") +
  scale_fill_manual(values = pal_overlay, name = "Treatment") +
  scale_x_continuous(breaks = c(0, 2, 4, 6)) +
  facet_wrap(~treatment) +
  labs(
    title = "Individual Haemoglobin Trajectories Over Time",
    x = "Month",
    y = "Haemoglobin (g/dL)"
  ) +
  theme_pubr() +
  theme(legend.position = "bottom")
```






## Mean per trajectories

```{r}
#| label: fig-meantraj
#| fig-cap: "Mean haemoglobin trajectory by treatment group with 95% confidence intervals."


pal_trt <- c("Control" = "black", "eSMC" = "darkgreen")
df_means2 <- df |>
  group_by(treatment, month) |>
  summarise(
    mean_hb = mean(hemoglobin),
    se_hb = sd(hemoglobin) / sqrt(n()),
    .groups = "drop"
  )

ggplot(df_means2, aes(x = month, y = mean_hb, colour = treatment, fill = treatment)) +
  geom_ribbon(
    aes(ymin = mean_hb - 1.96 * se_hb, ymax = mean_hb + 1.96 * se_hb),
    alpha = 0.2,
    colour = NA
  ) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  scale_colour_manual(values = pal_trt) +
  scale_fill_manual(values = pal_trt) +
  scale_x_continuous(breaks = c(0, 2, 4, 6)) +
  labs(x = "Month", y = "Mean haemoglobin (g/dL)") +
  theme_minimal()
```

# Model Building & Selection

## Linear Mixed Model {#sec-lmm}

### Model Rationale


### Unconditional Means Model (Base Model)

The first step in the modelling process was to fit an unconditional means model, also called the null model, which contains no predictors but includes a random intercept for each child.

$$\text{hemoglobin}*{ij} = \beta_0 + u*{0i} + \epsilon_{ij}$$



$$ICC = \frac{\sigma^2_{child}}{\sigma^2_{child} + \sigma^2_{residual}}$$

$$\text{~ indicates include only intercept}$$


```{r null-model-icc}
m0 <- lmer(hemoglobin ~ 1 + (1 | child_id), data = df, REML = TRUE)
summary(m0)
vc <- as.data.frame(VarCorr(m0))
var_child <- vc$vcov[vc$grp == "child_id" & vc$var1 == "(Intercept)"]

var_child

var_resid <- vc$vcov[vc$grp == "Residual"]
ICC <- var_child / (var_child + var_resid)
ICC
```

```{r}
null_fixed <- data.frame(
  Parameter = "Intercept",
  Estimate = round(fixef(m0)[1], 4)
)

null_random <- data.frame(
  Group = c("child_id", "Residual", "ICC"),
  Component = c("Intercept variance", "Residual variance", "Intraclass Correlation Coefficient"),
  Value = c(round(var_child, 4), round(var_resid, 4), round(ICC, 4))
)

null_fixed %>%
  kable(
    caption = "",
    format = "html",
    booktabs = TRUE,
    align = "c"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = FALSE, font_size = 14, position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#003366")
```

```{r}
null_random %>%
  kable(
    caption = "Random Effects and ICC from the Null Linear Mixed Model",
    format = "html",
    booktabs = TRUE,
    align = "c"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = FALSE, font_size = 14, position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#003366")
```


### Addition of Fixed Effects

After establishing the baseline variance structure, we introduced relevant predictors as fixed effects: month, treatment, age at baseline, and ITN use.






```{r fixed-effect}
m1 <- lmer(
  hemoglobin ~ month + treatment + age_baseline + itn_use + (1 | child_id),
  data = df,
  REML = FALSE
)
```

```{r}
tidy_m1 <- tidy(m1, effects = "fixed") |>
  mutate(across(where(is.numeric), ~ round(.x, 4)))

tidy_m1 %>%
  kable(
    caption = "Fixed Effects for the Model without Interaction",
    format = "html",
    booktabs = TRUE,
    align = "c"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = FALSE, font_size = 14, position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#003366")
```

### Addition of Interaction Term

To address the primary research question, an interaction between treatment and month was introduced.



**Same as**

```{r}
m2 <- lmer(
  hemoglobin ~ month + treatment + month:treatment + age_baseline + itn_use + (1 | child_id),
  data = df,
  REML = FALSE
)
summary(m2)
```




```{r}
m2 <- lmer(
  hemoglobin ~ month * treatment + age_baseline + itn_use + (1 | child_id),
  data = df,
  REML = FALSE
)
summary(m2)
```

```{r}
tidy_m2 <- tidy(m2, effects = "fixed") |>
  mutate(across(where(is.numeric), ~ round(.x, 4)))

tidy_m2 %>%
  kable(
    caption = "Fixed Effects for the Interaction Model",
    format = "html",
    booktabs = TRUE,
    align = "c"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = FALSE, font_size = 14, position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#003366")
```

```{r}
anova(m1,m2)
```







```{r interaction}
comp_table <- data.frame(
  Model = c("Model without interaction", "Model with interaction"),
  AIC = c(AIC(m1), AIC(m2)),
  BIC = c(BIC(m1), BIC(m2)),
  logLik = c(as.numeric(logLik(m1)), as.numeric(logLik(m2)))
)

lrt12 <- anova(m1, m2)

comp_table <- comp_table |>
  mutate(
    LRT_ChiSq = c(NA, round(lrt12$Chisq[2], 2)),
    Df = c(NA, lrt12$Df[2]),
    p_value = c(NA, signif(lrt12$`Pr(>Chisq)`[2], 4))
  )

comp_table %>%
  kable(
    caption = "Comparison of models with and without the treatment-by-time interaction",
    format = "html",
    booktabs = TRUE,
    align = "c",
    digits = 2
  ) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = FALSE, font_size = 14, position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#003366")
```




### Random Effects

The initial models assumed a random intercept only:

$$(1 \mid child_id)$$

We then added a random slope for month:

$$(1 + month \mid child_id)$$

```{r modelSlope}
m3 <- lmer(
  hemoglobin ~ month * treatment + age_baseline + itn_use + (month | child_id),
  data = df,
  REML = FALSE
)
```

```{r modelcomp_random}
comp_random <- data.frame(
  Model = c("Random intercept model", "Random intercept + random slope model"),
  AIC = c(AIC(m2), AIC(m3)),
  BIC = c(BIC(m2), BIC(m3)),
  logLik = c(as.numeric(logLik(m2)), as.numeric(logLik(m3)))
)

lrt23 <- anova(m2, m3)

comp_random <- comp_random |>
  mutate(
    LRT_ChiSq = c(NA, round(lrt23$Chisq[2], 2)),
    Df = c(NA, lrt23$Df[2]),
    p_value = c(NA, signif(lrt23$`Pr(>Chisq)`[2], 4))
  )

comp_random %>%
  kable(
    caption = "Comparison of random intercept and random slope models",
    format = "html",
    booktabs = TRUE,
    align = "c",
    digits = 2
  ) %>%
  kable_styling(bootstrap_options = c("striped", "bordered"),
                full_width = FALSE, font_size = 14, position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#003366")
```



```{r lmm-fit}
model_main <- lmer(
  hemoglobin ~ treatment * month + age_baseline + itn_use + (month | child_id),
  data = df,
  REML = TRUE
)
```

```{r}
tbl_fixed <- tidy(model_main, conf.int = TRUE, effects = "fixed") |>
  mutate(
    term = recode(term,
      "(Intercept)" = "Intercept",
      "treatmenteSMC" = "Treatment (eSMC vs Control)",
      "month" = "Month",
      "age_baseline" = "Age at baseline",
      "itn_useYes" = "ITN use (Yes vs No)",
      "treatmenteSMC:month" = "Treatment × Month"
    ),
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3),
    p.value = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value)),
    `95% CI` = paste0(conf.low, " to ", conf.high)
  ) |>
  select(term, estimate, std.error, `95% CI`, p.value)

tbl_fixed |>
  kable(
    caption = "Final mixed model fixed effects",
    col.names = c("Predictor", "β", "SE", "95% CI", "p-value"),
    format = "html",
    align = "c"
  ) |>
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "bordered", "condensed"),
    font_size = 14
  ) |>
  row_spec(0, bold = TRUE, color = "white", background = "#003366")
```

```{r icc}
icc_val <- icc(model_main)
r2_val  <- r2(model_main)

cat("--- Intraclass Correlation Coefficient (ICC) ---\n")
print(icc_val)
cat("\n--- Model R² (marginal and conditional) ---\n")
print(r2_val)
```

# Model Diagnostics {#sec-diagnostics}

```{r diagQQ}
fitted_vals <- fitted(m3)
resid_vals  <- resid(m3)

re <- ranef(m3)$child_id

par(mfrow = c(2, 2))

plot(fitted_vals, resid_vals,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, lty = 2)

qqnorm(resid_vals, main = "Q-Q Plot of Residuals")
qqline(resid_vals)

qqnorm(re[, "(Intercept)"], main = "Q-Q Plot of Random Intercepts")
qqline(re[, "(Intercept)"])

qqnorm(re[, "month"], main = "Q-Q Plot of Random Slopes")
qqline(re[, "month"])

par(mfrow = c(1, 1))
```


